\thetitle

### 2021-02-09 writing/reviews/pldi-rebuttal.md

PLDI 2021 Paper #66 Author Response
===========================================================================

First of all, we thank all of the reviewers for reading our submission, asking
critical questions, and offering constructive feedback. Our responses to all
comments and questions follow, in which we quote each comment/question and then
respond below.

Review #66A
===========================================================================
> - Could you comment a bit more on what makes a mutator "good"? (other than
>   experimentally seeing that it is good.) Perhaps, what would be a really bad
>   mutator for evaluating blame assignment?

A bad mutator results in a type-level bug that is trivial to locate. For
instance, a bug that leads to a type-level error during the evaluation of the
component that contains it is trivial to locate. In contrast, a bug that is
non-trivial to locate is one that causes the buggy component to evaluate without
an error but produce instead a value of the wrong type. An interesting bug is
one where the value with the wrong type manages also to cross undetected to at
least another component before a type-level check flags it. Good mutators should
be able to produce these interesting bugs.

@@

> - L956: 30,000 hours of compute time is 1,250 days. I assume that the CPU has
>   many cores. Could you put that number?

Sure. Each CPU has 28 doubly-threaded cores.

> - L991: Figure 6: It took me some time to understand the figure. I wonder if
>   there is a better way to represent the same data?

We have reached the same conclusion and we are considering alternatives that
will make it easier to see how two modes compare to each other, such as a series
of two-sided bar charts.

@@

Review #66B
===========================================================================
> The experiment results show that there are debugging scenarios where
> Transient is more useful than Natural. Does it happen simply because
> of luck? Or, does it imply that Transient is more suitable than
> Natural for a particular kind of bugs for some reason? Also, can there
> be a new blame strategy that outperforms both Transient and Natural
> all the time?

We were also surprised to discover that Transient outperforms Natural in some
scenarios. After all, the theory of blame says otherwise. Even though we haven't
identified specific patterns, we believe that the reason behind Transient's
unexpected success is due to the advanced features of Typed Racket's type system
that haven't appeared in models that compare Natural and Transient:
polymorphism, row types, mixins, and more.

We do not have any clues about a strategy that outperforms both Natural and
Erasure. That said, our current conjecture is that a simple strategy that just
points somewhere in the blame trail may be good enough. Such a strategy would
also enable optimizations that Typed Racket currently foregoes in order to track
blame. The good thing is that now we have a method to compare different
strategies.

@@

chrdimo: TODO We need more details from Ben here.

> Theoretical gradual type systems usually allow programmers to control
> the precision of type annotations in a fine-grained way with the Dyn
> type rather than distinguishing typed and untyped modules. Would the
> same evaluation method be able to directly appliable to blame in such systems?

Yes. Incorporating type Dyn boils down to considering a hierarchy of types for
each component rather than a single type. At the same time, our method is
parametric to the "size" of a component and adapts in a straightforward manner
to a setting where each definition can be considered a component instead of a
whole module. Of course both of these adjustments lead to significantly larger
scenario lattices than the ones we investigate, so keeping the experiment
computationally feasible would require more aggressive sampling.

@@

> Please explain what impedance mismatches are in the paper. Is it
> simply type mismatches?

An impedance mismatch denotes a type annotation that is detected to disagree
with the actual type of the annotated code. We will adjust the prose to clarify.

@@

> The paper states that the effort distribution of the random mode
> follows a normal distribution, as expected. Please explain why a
> normal distribution is a reasonable expectation. ...
> This point does not affect the contribution of the paper, but it
> is not clear from the paper.

You are right that for a given trail of length $n$ the probability that the
random rational programmer discovers the buggy component on every step is $1/n$.
The missing piece is that we randomly sample from the set of scenarios -- and by
extension the set of trails. So the shape of the random programmer's effort
matches our sampling distribution. That said, as you observe, the distribution
of the random programmer's effort is not relevant to our results, so we will
remove the characterization.

### 2021-02-09 writing/reviews/pldi-rebuttal-notes.md

chrdimo? We need more details from Ben here.

Theoretical gradual type systems usually allow programmers to control
the precision of type annotations in a fine-grained way with the Dyn
type rather than distinguishing typed and untyped modules. Would the
same evaluation method be able to directly appliable to blame in such systems?

  addressed
  TODO: The answer is yes, but this needs more thought to frame the answer the right way.
  ⟶ Our evaluation framework should be directly applicable to blame in gradual typing systems using Dyn.
    The details of how it applies depends on the granularity of the blame information offered by the system -- in other words, what the system considers components.
	If components are modules, as in Typed Racket, then nothing changes.
	Systems with Dyn more likely treat top-level definitions as components, in which case definitions are individually typed (the whole definition is annotated) rather than modules.

chrdimo: Yes. Incorporating type Dyn boils down to considering a hierarchy
of types for each component rather than a single type. At the same time,
our method is parametric to the ``size'' of a component and adapts in a
straightforward manner to a setting where each definition can be
considered a component instead of a whole module. Of course both of these
adjustments lead to significantly larger scenario lattices than the ones we
investigate and they have to be sampled aggressively and carefully.

@@

Please explain what impedance mismatches are in the paper. Is it
simply type mismatches?

  ⟶ Impedance mismatches are mismatches between a piece of code's true type and its annotation.
    For instance, the code `(λ (x) (+ x 42))` has type `Number -> Number` but might be annotated as `Number -> String`, causing an impedance mismatch.
    We will adjust the text to clarify or remove this terminology.

chrdimo: An impedance mismatch denotes a type annotation that doesn't
match the actual type of the annotated code. We will adjust the prose to
clarify.

@@

# Minor comments

The paper states that the effort distribution of the random mode
follows a normal distribution, as expected. Please explain why a
normal distribution is a reasonable expectation. When there are $n$
untyped components, the random mode randomly selects one of them, and
only one of them is a buggy component. Therefore, in the first
iteration, the random mode reaches the end of the trail with the
probability of $1/n$. In the second iteration, there are $(n-1)$
untyped components and the probability of a failure at the first
iteration is $(n-1)/n$, so the probability of effort $2$ is $1/n$
again. In this way, it seems to be that the probability of effort $i$
is $1/n$ where $1<=i<=n$ and $n$ is the number of initially untyped
components. Each trail has a different number of untyped components,
so the resulting distribution would be the average of such
distributions. Based on this reasoning, it is unclear why the
distribution is a normal distribution. This point does not affect the
contribution of the paper, but it is not clear from the paper.

  addressed
  TODO: response sketch: you're right, we misspoke, this is not the normal distribution; our reasoning for why the normal-ish shape of distribution we see makes sense is that given a fixed trail the probability at each step of the trail ending is 1/n, but we randomly select trails so we're multiplying the random distribution of trail lengths by that 1/n one? Not clear on the details of this reasoning.
  ⟶

chrdimo: You are right that for a given trail of length n the probability
that the random rational programmer discovers the buggy component after k 
attempts is always 1/n and independent of k. However we sample randomly
from the set of scenarios and as a result the set of trails. So the curve 
of the random programmer matches our sampling distribution. That said, as
you observe, what is the distribution of the random programmer is an
irrelevant point to our results and we will remove the characterization.

@@

Comments for author
-------------------
This is a neat paper that develops an approach for evaluating gradual typing and blame assignment. Thanks for submitting it to PLDI '21! The results are presented in an extremely clear way that draws out the key insights without compromising on technical detail. The notion of a rational programmer -- loosely modeled on rational agents from economic theory -- is a neat conceptual device. And the empirical approach does seem like the right way to address the question of the usefulness of blame.

But while there is a lot to like about the paper, it also has some shortcomings. The suite of programs used for the evaluation is fairly small (c.f., the SIGPLAN checklist).

  addressed
  TODO: respond to this ^, saying that we use the standard suite of benchmarks for evaluating GT -- point to Ben's POPL paper and others, and we use all of the largest benchmarks in the suite

chrdimo: Could you please clarify what you mean by fairly small with
respect to the SIGPLAN checklist? We are the first to conduct an empirical
study for blame and gradual typing so we repurposed the standard
performance benchmark suite for Typed Racket and, in particular, its
largest and most complex benchmarks. Our inspection of the benchmarks
indicates that they are representative set of Racket programs in terms of
features and structure.

@@

The faults are introduced using synthetic mutations that may or may not correspond to the kinds of errors that arise in practice. This can be seen from the results in Figure 7 -- the lengths of the "trails" is quite small.

  addressed
  TODO: unclear that syntheticness relates to trail length at all, and synthetic bugs are the only option here -- there is no corpus of documented GT bugs in the wild anywhere. Indeed, most type errors never even make it into code repositories. Note that we are the first to evaluate blame in this setting.

chrdimo: There is no existing catalogue of bugs in gradually typed
programs. Furthermore, most type-level bugs do not survive after decent testing and 
thus do not make it to code repos. There are a few anecdotes from the literature but they are mostly
artificial. Hence, since we need a large number of buggy scenarios for a
meaningful study, a synthetic approach is the only way forward. In
addition, with our mutation we can tune the set of scenarios to get a
diverse set of bugs.

@@

While the study is mostly well done, it only considers two blame assignment strategies.
For example, it would be interesting to consider the "omniscient programmer" to get a bound on the value of blame. But this is not done.

  addressed
  TODO: We don't understand what is meant by "omniscient programmer", but our best understanding is the ideal oracle strategy that always picks the right module right away and thus always finds the bug in one step. We didn't show this because it's trivial. We do have other baselines to understand blame's performance: the random mode, and the exceptions modes for each of the semantics.

chrdimo: Could you please explain what you mean by "omniscient
programmer"? If this is the programmer that always makes the best choice,
i.e., finds the bug at the first try then this programmer is trivial. It
has 100% success rate and a constant blame trail of 1. Thus comparing it
with any strategy amounts to the number of scenarios where a strategy
fails and those where it succeeds but the blame trail has length greater
than one. This information is in the paper but If you think it would be
useful, we point it out. Moreover, the experiment considers 10 different
modes of the rational programmer spread across 3 different gradual typing
systems.

@@

And unfortunately the results are inconclusive -- Section 8 says "our results call for a deeper understanding of the two models for blame".

  addressed
  TODO: This is what makes our results surprising: they challenge the theoretical conclusions that Natural is superior to Transient, suggesting that the theory is not good enough to understand the pragmatics of blame in reality.

chrdimo: The quote from section 8 refers to the theory of blame rather
than our results. Our results show in clear terms that, at least for our
benchmarks, a blame strategy is superior that exceptions and that natural
and transient have comparable efficacy. The latter is surprising given that
theoretical models (and our initial expectations) point to the opposite.
Thus a side-effect of our experiment is a call to theoreticians for models
that are attuned with practice.

@@

Also, the published Transient scheme seems to have a minor bug, which might affect the outcomes when fixed (Section 8.2).
  addressed
  TODO: 8.2 does not indicate a bug in Transient, but rather a defect in the design of the system. ... something about a gap between Reticulated Python and

chrdimo: Section 8.2 does not discuss a bug in Transient but rather a
limitation of its current design that we discovered exactly because of the
experiment.

@@

Most if not all of these shortcomings are discussed in the paper, which is much appreciated. But they do risk undermine the value of the contributions to some degree.

chrdimo: We would like to point out that the results of the comparison
between the tree systems are only one part of the contribution of the
paper.  The other part is the method itself. Benchmarks, source of
interesting bugs and rational programmer modes are all parameters that can
be instantiated differently without affecting the recipe we discuss at the
end of section 2 nor the analytical framework from section 5 that
underlies our experiment.

@@

To explain one comment above: I was surprised not to see a comparison against an "omniscient programmer" to establish a baseline (c.f., the SIGPLAN checklist). That is, model a blame assignment scheme in which blame could be placed on any component. And always pick the component that minimizes the length of the trail. Of course, this would be infeasible to implement in practice, but it would place an upper bound on the utility of blame.

chrdimo: Our experiment does have a baseline. The random mode of the
rational programmer serves as the ``control'' of our experiment. 
Furthermore the exception modes serve as further ``controls'' to
distinguish the effect fo checks from that of blame,

Similarly, I wondered why the modes of the transient programmer only consider picking the first/last blame, and not other choices. Section 5.2 just says "our answer" is that there are at least "two reasonable options." Why are other choices not reasonable?
  addressed
  TODO: Answer suggested by Ben: it's unspecified how one should interpret Transient blame, and there isn't time to evaluate all possible choices, so we pick the most reasonable interpretations.

chrdimo: The transient semantics provide no interpretation for the blame
set. We picked the first and last because they match the intuitive
interpretation of the blame set as a list. We didn't mean to imply that
other options are unreasonable. We just needed to make a reasonable choice 
and keep the duration of the experiment within reasonable limits.

@@

The current paper extends the POPL 2020 methodology from contracts in
Racket to gradual (or migratory) typing in Typed Racket, in several
steps:

First, to complement the "Natural" higher-order contract system of
(Typed) Racket, this work implements a "Transient" blame assignment
strategy, which unlike Natural inserts only shallow run-time checks.
(To date, Transient has appeared only in the separate design and
implementation of Reticulated Python, which prevents a head-to-head
comparison as desired in this work.)

Second, compared to POPL 2020, new mutators are defined to inject the
kinds of type-mismatch bugs that may be caught by the gradual type
system and corresponding run-time checks and blame assignment.

Third, the paper runs a large experiment on 10 benchmark programs,
inserting mutations and analyzing whether the blame trails produced by
each debugging mode --- following (a) the component blamed by Natural,
(b/c) the first or last component, respectively, blamed by Transient,
or (d) an exception raised by the underling language --- conclude
with the actual bug. Results show that Natural, Transient-First, and
Transient-Last all produce short successful blame trails (typically of
length one or two) on most debugging tasks, and that they are all
similarly effective with no clear winner for all tasks.

@@

1. That Natural is not much more effective than Transient. I thought
the shallow checks of Transient would certainly be less effective than
the full checks of Natural. It will be interesting to see if and how
this holds up in subsequent usability studies, both with synthetic and
real users and tasks.

2. That Transient --- as implemented in Typed Racket --- is much
slower than Natural. This also goes against intuition, as well as
previous results reported about Reticulated Python (but apparently
there were flaws in those experiments, as noted in this paper). Given
that the experiments are developed and run in the context of the
large, real-world Racket ecosystem, I wonder whether about the chance
of incidental implementation choices or bugs affecting this result.

  TODO: We should clarify that checks alone is faster, but adding blame is slower (bc blame map is linear).

chrdimo: A side note here: Transient Typed Racket is faster than Natural
Typed Racket when we turn off blame. Exactly as a sanity check that this is not 
an artifact of our implementation we have coded up benchmarks in Python
and we confirmed that in Reticulated the performance of transient with blame is 
even worse than what we have observed in Typed Racket.

@@

This seems to be "only" an engineering challenge.

And the presentation in this section is rather informal. As someone
with passable knowledge of Natural and Transient, I would have liked
to see a concise formalism with clear "knobs" for choosing between
blame assignment strategies. Trying to imagine readers with less
familiarity, I wonder how well this discussion serves as a primer.

  TODO: this is a minor part of the paper (not a result), there is a large gap between the model and what's needed to scale to a full lang -- it's not clear from the simple model how to design check insertion and blame tracking for new features. This required research, will be published elsewhere

chrdimo:

We disagree that implementing Transient Typed Racket is mere engineering.
Typed Racket comes with a way more sophisticated type system than
Reticulated and there are a lot of points around type checks and blame
that neither Reticulated nor the Transient model provide answers for.  We
are happy to share with the chair a recent dissertation chapter that
describes all the ideas and work that went into Transient Typed Racket.

We believe that adding a formal model to the paper would distract
from the main contribution of the paper as there are already multiple
publications on the semantics of Transient, Natural and Erasure and at
least one that compares them formally [Greenman et al. OOPSLA 2019]. That said, we
agree that we should improve the presentation in section 3 to introduce
more context.

@@

* `arithmetic`: Simpler version of previous
* `boolean`: Same as in previous
* `negate-cond`: Same as in previous

So there is quite a bit of similarity. And on the flip side, why are
not all (8) mutators from the POPL 2020 experiment included here?

These questions are not so important in an absolute sense; _some_
specific mutators need to be chosen to generate some bugs, and these
seem to work fine. But they are salient given how much the paper
emphasizes the new mutators as a contribution.

chrdimo: While we agree that there is overlap between our mutators
and those from Lazarek's POPL work (and the common set of mutators where
the latter come from), we disagree that the construction of custom
mutators is straightforward. The sixteen we include in the paper are 
not the original set we tried and others that seemed like good candidates
have proven to be ineffective. For instance replacing car with cdr should
lead to type-level bugs however none of them are interesting as we define
interesting in the paper. Similarly, swapping mutable vectors and hashes
with immutable ones is an interesting mutation in the traditional sense
and should lead to a type error in Typed Racket as its type system
distinguishes the two, however we discovered that again is not interesting
in our setting. In general, there is no methodology for constructing a new
set of mutators, so narrowing down the ones in the paper required
that we first characterize precisely what makes a bug interesting and then 
rounds of trials to end up with a set of mutators that leads to a 
sufficiently large set of interesting diverse bugs.

@@

  TODO: the analytical part is important: we described all of these things in a common framework, allowing for comparison

I'm also not sure why the exception mode is needed in addition to
Erasure (I was confused in Section 2 L204-215 and Section 5.1 L732).
Indeed, Section 5.3 defines the Erasure mode the follow the Natural
exceptional mode.

  TODO: the exception modes are a key part of the analysis framework, enabling understanding of the value of blame separately from checking

chrdimo:

We disagree that section 5 boils down to picking the first and last modes
for Transient. The core of this section is the analysis of of the
  different systems and their unification in a common precise framework as
  modes. Because of the common substrate of blame trails modes enable
  the comparison between the different systems even when sampling is
  involved.  Furthermore we use modes to introduce in a uniform manner in
  the framework baselines that help us isolate confounding factors from
  the actual effectiveness of blame. This si exactly the role of the
  exception modes for Natural and Transient. Those two perform checks at
  runtime and the exception modes help us determine what part of the
  success of Natural and Transient is due to the extra checks or due to
  blame (and also whether blame masks useful information from exceptions).
  For example, we see that Natural without blame but with the same checks
  outperforms Erasure in ~25% of the scenarios (figure 6). Blame add
  another ~11% on top of that. For Transient there is a small percentage
  of scenarios where exceptions due better than blame but still blame
  improves over exceptions by ~11% compared to Erasure.

chrdimo: As a final note, even though each of the above points
individually may not have sufficient technical depth for a separate
contribution (which we do not claim), together they exhibit the creative
steps that are necessary to lift the scope and magnitude of the idea
from Lazarek's POPL paper on evaluating blame from a single behavioral
contract system to a framework for evaluating and comparing different
gradual typing systems.

### 2021-02-09 writing/reviews/pldi-rebuttal-notes.md

Comments for author
-------------------
This work advances a compelling approach to evaluating a usability
question through a systematic exploration of synthetic bugs and fixes.
The large-scale experiment is impressive, and I find both of the main
results surprising:

1. That Natural is not much more effective than Transient. I thought
the shallow checks of Transient would certainly be less effective than
the full checks of Natural. It will be interesting to see if and how
this holds up in subsequent usability studies, both with synthetic and
real users and tasks.

2. That Transient --- as implemented in Typed Racket --- is much
slower than Natural. This also goes against intuition, as well as
previous results reported about Reticulated Python (but apparently
there were flaws in those experiments, as noted in this paper). Given
that the experiments are developed and run in the context of the
large, real-world Racket ecosystem, I wonder whether about the chance
of incidental implementation choices or bugs affecting this result.

  addressed
  TODO: We should clarify that checks alone is faster, but adding blame is slower (bc blame map is linear).

@@

*Natural and Transient Under the Same Roof (Section 3)*

This seems to be "only" an engineering challenge.

And the presentation in this section is rather informal. As someone
with passable knowledge of Natural and Transient, I would have liked
to see a concise formalism with clear "knobs" for choosing between
blame assignment strategies. Trying to imagine readers with less
familiarity, I wonder how well this discussion serves as a primer.

  addressed
  TODO: this is a minor part of the paper (not a result), there is a large gap between the model and what's needed to scale to a full lang -- it's not clear from the simple model how to design check insertion and blame tracking for new features. This required research, will be published elsewhere

@@

chrdimo: While we agree that there is overlap between our mutators
and those from Lazarek's POPL work (and the common set of mutators where
the latter come from), we disagree that the construction of custom
mutators is straightforward. The sixteen we include in the paper are 
not the original set we tried and others that seemed like good candidates
have proven to be ineffective. For instance replacing car with cdr should
lead to type-level bugs however none of them are interesting as we define
interesting in the paper. Similarly, swapping mutable vectors and hashes
with immutable ones is an interesting mutation in the traditional sense
and should lead to a type error in Typed Racket as its type system
distinguishes the two, however we discovered that again is not interesting
in our setting. In general, there is no methodology for constructing a new
set of mutators, so narrowing down the ones in the paper required
that we first characterize precisely what makes a bug interesting and then 
rounds of trials to end up with a set of mutators that leads to a 
sufficiently large set of interesting diverse bugs.

*Debugging Strategies (Section 5)*

These also seem to be overemphasized. The main design choice seems to
be in how to use Transient blame, which reports multiple components,
and picking the first and last are pretty intuitive.

  addressed
  TODO: the analytical part is important: we described all of these things in a common framework, allowing for comparison

@@

I'm also not sure why the exception mode is needed in addition to
Erasure (I was confused in Section 2 L204-215 and Section 5.1 L732).
Indeed, Section 5.3 defines the Erasure mode the follow the Natural
exceptional mode.

  addressed
  TODO: the exception modes are a key part of the analysis framework, enabling understanding of the value of blame separately from checking

### 2021-02-09 writing/reviews/pldi-rebuttal.md

> Similarly, I wondered why the modes of the transient programmer only consider picking the first/last blame, and not other choices. Section 5.2 just says "our answer" is that there are at least "two reasonable options." Why are other choices not reasonable?

The transient semantics provide no interpretation for the blame set. We picked
the first and last because they match the intuitive interpretation of the blame
set as a list. We didn't mean to imply that other options are unreasonable. We
just needed to make a reasonable choice and keep the duration of the experiment
within reasonable limits. We will clarify this point in the prose.

Review #66D
===========================================================================

> Given that the experiments are developed and run in the context of the
> large, real-world Racket ecosystem, I wonder whether about the chance
> of incidental implementation choices or bugs affecting this result.

A side note here: Transient Typed Racket is faster than Natural Typed Racket
when we turn off blame. Exactly as a sanity check that this is not an artifact
of our implementation we have coded up benchmarks in Python and we confirmed
that in Reticulated the performance of transient with blame is even worse than
what we have observed in Typed Racket.

> *Natural and Transient Under the Same Roof (Section 3)*
>
> This seems to be "only" an engineering challenge.

We disagree that implementing Transient Typed Racket is mere engineering.
Typed Racket comes with a way more sophisticated type system than
Reticulated and there are a lot of points around type checks and blame
that neither Reticulated nor the Transient model provide answers for.  We
are happy to share with the chair a recent dissertation chapter that
describes all the ideas and work that went into Transient Typed Racket.

We believe that adding a formal model to the paper would distract
from the main contribution of the paper as there are already multiple
publications on the semantics of Transient, Natural and Erasure and at
least one that compares them formally [Greenman et al. OOPSLA 2019]. That said, we
agree that we should improve the presentation in section 3 to introduce
more context.

> *Custom Mutators (Section 4)*
> The paper suggests that the mutators required here are significantly
> different and more subtle, but this does not seem to be the case when
> comparing Figure 3 and Table 3 of the POPL 2020 paper.
> ...

While we agree that there is overlap between our mutators and those from
Lazarek's POPL work (and the common set of mutators where the latter come from),
we disagree that the construction of custom mutators is straightforward. The
sixteen we include in the paper are not the original set we tried and others
that seemed like good candidates have proven to be ineffective. For instance
replacing car with cdr should lead to type-level bugs however none of them are
interesting as we define interesting in the paper. Similarly, swapping mutable
vectors and hashes with immutable ones is an interesting mutation in the
traditional sense and should lead to a type error in Typed Racket as its type
system distinguishes the two, however we discovered that again is not
interesting in our setting. In general, there is no methodology for constructing
a new set of mutators, so narrowing down the ones in the paper required that we
first characterize precisely what makes a bug interesting and then rounds of
trials to end up with a set of mutators that leads to a sufficiently large set
of interesting diverse bugs.

> *Debugging Strategies (Section 5)*
>
> These also seem to be overemphasized. The main design choice seems to
> be in how to use Transient blame, which reports multiple components,
> and picking the first and last are pretty intuitive.
>
> I'm also not sure why the exception mode is needed in addition to
> Erasure (I was confused in Section 2 L204-215 and Section 5.1 L732).
> Indeed, Section 5.3 defines the Erasure mode the follow the Natural
> exceptional mode.

We disagree that section 5 boils down to picking the first and last modes for
Transient. The core of this section is the analysis of of the different systems
and their unification in a common precise framework as modes. Because of the
common substrate of blame trails modes enable the comparison between the
different systems even when sampling is involved. Furthermore we use modes to
introduce in a uniform manner in the framework baselines that help us isolate
confounding factors from the actual effectiveness of blame. This si exactly the
role of the exception modes for Natural and Transient. Those two perform checks
at runtime and the exception modes help us determine what part of the success of
Natural and Transient is due to the extra checks or due to blame (and also
whether blame masks useful information from exceptions). For example, we see
that Natural without blame but with the same checks outperforms Erasure in ~25%
of the scenarios (figure 6). Blame add another ~11% on top of that. For
Transient there is a small percentage of scenarios where exceptions due better
than blame but still blame improves over exceptions by ~11% compared to Erasure.

As a final note, even though each of the above points individually may not have
sufficient technical depth for a separate contribution (which we do not claim),
together they exhibit the creative steps that are necessary to lift the scope
and magnitude of the idea from Lazarek's POPL paper on evaluating blame from a
single behavioral contract system to a framework for evaluating and comparing
different gradual typing systems.

> *Additional Comments and Typos*
>
> L69: I wish here, and elsewhere, the comparison to [13] had been made
> more explicit. As discussed above, there seem to be more similarities
> than suggested.
> L92: "use of _a_ higher-order contract system"
> L155: "authors extensive"
> L253: "exports _it_ as"
> L280: Missing `untyped-`
> L314: " . "
> L382: "Dyn,"
> Fig 2: Presumably authors were included to indicate the variety of
> sources, but I'm not sure that was necessary.
> L597: "make us_e_ of"
> L664: "described in _Section_ 4"
> L980: "Figure 5.4" ==> "Figure 6"
> L986: "2.8%" ==> "2.18%" ?
> L1017: "33%" ==> "26%" ?
> L1304: "Problem is ...."

Thank you. We will fix these.

@@

Review #66E
===========================================================================

> This paper seeks to establish (or refute) the benefit of blame tracking in gradually typed programming environments. It takes several Typed Racket benchmarks and considers multiple approaches to blame tracking. The central idea is to use a fault injection method to create variant programs along two dimensions: type errors inserted according to some crafted heuristics, and static typing selectively removed on a component-by-component granularity. This creates a large number of variant programs which may produce run-time type errors with blame; this is then iterated to simulate a "rational programmer", i.e. such that blame is used to select where more static checking will be added next; at any given iteration this may or may not catch the injected error, and the number of iterations is counted as a proxy for effort. A further contribution is implementing the different forms of blame-tracking in the same framework so that they can be compared. The finding is that blame is found to offer some positive advantage in effort (over simply being guided by the stack trace of a run-time exception) in 9--13% of the initial sample of variants, with some minor differences between the forms of blame tracking tried.

@@

  TODO: misunderstanding: we propose a methodology

The primary intended contribution of this paper is the methodology, and the
experimental results serve as secondary contributions that validate the
methodology's utility. We will rephrase the introduction to further emphasize
that focus. Furthermore, we would like to emphasize that the two key findings in
the experimental results are that 1) both Natural and Transient's dynamic type
enforcement with blame tracking improve significantly over Erasure, and 2)
Natural's blame tracking is not significantly better than Transient's, despite
the theoretical indications to the contrary.

> However, I don't find the results to be as conclusive as the authors state. Aside from the Erasure case, whose relevance I found unclear (see below), conclusions rest on some small-valued "more useful than" percentages. These are pretty hard to interpret, because the "usefulness" metrics necessarily build in a lot of simplifying assumptions. E.g. in 5.4, the "percentage of scenarios where ... is more effective" doesn't account for *extent* of difference, and the trail-length "programmer effort" metric in 5.5 is also unlikely to be a great proxy. These limitations are understandable, method-wise, but overall the results are fairly null... they let us continue believing "what we'd expect" but not with any great sense of added confidence. That doesn't entirely diminish the work, of course, but I'd say the current write-up overplays its hand a little. I appreciate the discussion of threats to validity.

  TODO: misunderstanding: the results are not at all "what we'd expect" based on the theory

The status quo of what we would expect based on the current theory of Natural
and Transient is that 1) either kind of blame tracking should be better than
none, and 2) Natural should provide significantly better blame information than
Transient and Erasure. Our results validate the first expectation that blame
provides benefits over plain exceptions. Surprisingly, our results contradict
the second expectation and suggest that in fact Natural and Transient offer
comparable blame information.

> About Erasure: if I'm reading this correctly, the key thing here is that in Typed Racket, the set of static types is more expressive than the set of dynamic types. For example, there is a static notion of non-negative integer that is distinct from plain integer. This sort of design isn't universal -- in some languages/systems static types closely mirror the classification of objects in the language's dynamic semantics. Writing these stricter contracts into a program brings its own benefits, separate from blame or indeed from static checking. It feels like the paper doesn't take enough care to distinguish the two effects: the effect of systematically applying more refined contracts over program values, and the effect of gradually enabling static checking of those contracts (iteratively, guided by blame). The authors do mention this around line 204, and the comparisons in Figure 6 between "_ exceptions" and "Erasure" seem to be measuring this -- the gain from checking these extra annotations, with the more precise and/or more timely checks that they imply, relative to the erased case where only the language baseline contracts are checked. Indeed that's the point of 'exception' experiments. Since the biggest effect sizes on display are these ones -- between Erasure and anything else -- this seems at best distracting. I'd be glad to hear from the authors if I'm misunderstanding anything here. In the detailed comments below, I have noted some places in the text where it would be useful to remind the reader that this design property of Typed Racket is at play.

TODO: misunderstanding about gradual typing as a whole, and the point of the exception modes

The idea of the Erasure semantics is not related to the expressiveness of the
type system or how the types relate to kinds of runtime values. Rather, the key
idea is that Erasure does not enforce type annotations with dynamic checks at
all. Running a gradual program using the Erasure semantics is the same as
stripping all annotations and running the resulting (completely dynamic)
program. Thus, Erasure programs can only raise exceptions from either the checks
performed by the language's primitive operations (e.g. `+`), or from the program
itself raising an exception. In contrast, the other two semantics do enforce
type annotations dynamically; both Natural and Transient insert dynamic checks
verifying that the shape of values at runtime match their static annotations.
Thus, in addition to exceptions, gradual programs using these semantics may
raise a second kind of error: a dynamic type check failure. These dynamic errors
can include blame information if the dynamic type checks are designed to track
blame, or they can forego blame and just offer a stack trace. In the case of
Natural, these two options are captured by the "Natural" and "Natural
exceptions" modes, respectively. Thus the "Erasure" mode represents a completely
different kind of checking (and thus error reporting) compared to both the
"Natural" and "Natural exceptions" modes (the two of which share the same kind
checking, but different error reporting).

> It's interesting to note that there seems to be no convincing analogous experimental or modelling-based justification for plain old static typing, i.e. showing that it somehow presents a net gain to the programmer. Rather, this has simply been posited/assumed by a very long line of work. I am not defending that state of affairs, but it points to the difficulty of showing conclusively that something truly helps programmers. I can see value in this sort of simulation-style approach, and it is no less convincing than user studies. So I'd be interested to hear from the author(s) if they have any more arguments that (1) I've underestimated the results' conclusiveness, or (2) this is a novel/interesting family of methods that might be pursued more widely, or (3) that deeper experiments building on these ideas might yield more compelling insights.

(Matthias) As posed, the question expresses a significant mis-characterization
of the submission. It is not the benefits of a gradual type system that are in
doubt. The experimental setup keeps the type system constant but allows to
answer the question which of several run-time checking regimes (for enforcing
type consistency) provides the best explanatory messages in case of violations.
Hence the analogous question for purely static type systems would ask which of
several reporting schemes provides the best explanatory message for type-errors.
In the case of simply typed languages and even language with local type
inference, this question is basically meaningless. In the case of languages with
HM type inference, the question has been implicitly raised for four decades with
the development of alternative ways of finding the source of inference
conflicts.

If the reviewer is indeed interested in the question of whether type systems
help programmers---a question that this submission does _not_ ask---the recent
OOPSLA literature contains several user studies. A simple Google query will
suggest a short list of these publications.

  TODO ll: I added responses (after ⟶) to anything below that seems to warrant more than "Thanks, we'll fix it".

### 2021-02-09 writing/reviews/pldi-rebuttal.md

PLDI 2021 Paper #66 Author Response
===========================================================================

Thank you to al reviewers for the helpful feedback. 
critical questions, and offering constructive feedback. Our rebuttal combines each comment/question with a response.
